---
title: "Best of Both Worlds: RStudio and Python"
author: "Manya Shah"
date: 2021-05-29T21:13:14-05:00
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Python or R?

Both languages have their pros and cons, and are better for serving some purposes than others. Python is more of a general language while R specialises in statistical analysis. Coding in python almost feels like texting, whereas R can have a slighly more complicated syntax. The reticulate library in RStudio can let us combine the two languages and help achieve a seamless coding experience. 

Lets try to run some basic commands in both R and python and see what happens.

```{r cars}
library(reticulate)
hi <- "Hello"
hi
```

```{python}
hi="World"
print(r.hi,hi)
```

```{r}
cat(c(hi,py$hi))
```

We were able to combine both the R and Python output to get our final answer "Hello World"

## What about transferring Python objects to R? 

Lets create two vectors in Python, and try to plot them using R. 

```{python}
import pandas
flights = pandas.read_csv("flights.csv)
```

```{r}
library(ggplot2)
ggplot(py$flights,aes(carrier,arr_delay))+geom_point()+geom_jitter()
```

